Fast Sorting on Modern Computers

### 2005-05-30 paper/docs/ordernsquared.tex
spell checked updateed intro read through method removed extras, which was part of the demo added labels
finished 4 chapters

Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
manner in which they check every key in the array in order to sort a single key;
for every key to be sorted, every other key is checked.  Four of these sorts are
discussed here. They are insertion sort, selection sort, bubblesort and
shakersort.

@@

The level 1 cache should have equally bad performance as a result. Since it is
smaller than the array, the performance of the level 1 cache should simulate the
performance of the level 2 cache.

@@

\section{Insertion sort}

@@

\section{Bubblesort2}
Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
implementation above ends early if it detects that the array is fully sorted.
It keeps a binary flag to indicate if there were any exchanges along the way.
Instead, bubblesort2 keeps track of the last exchange made, and starts the next
iteration from there. This means that any pockets of sorted keys are skipped
over, not just if they occur at the end. Bubblesort2 code is in Figure
\vref{Bubblesort2 code}

@@

\section{Shakersort}
Shakersort is sometimes called back-bubblesort. It behaves like a bubblesort
which moves in both directions rather than just one way. Both ends of the array
become sorted in this manner, with the larger keys slowly moving right and the
smaller keys moving left. Sample code for shakersort is in Figure
\vref{Shakersort code}.

\code{Shakersort code}{on2_shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys. In addition, the worst case for bubblesort is not a
worst case for shakersort and is less likely.

The cache performance of shakersort is expected to be superior to that of
bubblesort. A small amount of temporal reuse occurs at the ends of the array,
as each key is used twice once its loaded into the cache. In addition, when the
array is not more than twice the size of the cache, then the keys in the centre
of the array are not ejected from the array at all.

@@

% I'm doing a slightly different result here. Level 2 doesn't interest us, so put
% in 2 prediction results instead - 1 for select/insert, 1 for bubble/shaker
\afterpage{
\label{Elementary sort results}

@@

Insertion sort performs best of the four sorts in relation to cache misses. This
is due to the fact that the keys it considered most recently are used again for
the next key. These keys will be swapped out fairly regularly, however, but not
nearly as often as selection sort, where they are swapped out at nearly every
iteration.

>>>>

\label{ordernsquared}
Elementary sorts are also referred to as $O(N^2)$ sorts. This is due to the
manner in which they check every key in the array in order to sort a single key;
for every key to be sorted, every other key is checked.  Four of these sorts are
discussed here. They are insertion sort, selection sort, bubblesort and
shakersort.

@@

The level 1 cache should have bad performance as a result. Since it is smaller
than the array, the performance of the level 1 cache should simulate the
performance of the level 2 cache.

@@

%TODO look at knuths number here
TODO Knuth has a formula which I will look up.
\section{Insertion sort}

@@

\section{Bubblesort2}
Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
implementation described in the previous section ends early if it detects that
the array is fully sorted.  It keeps a binary flag to indicate if there were any
exchanges along the way.  Instead, bubblesort2 keeps track of the last exchange
made, and starts the next iteration from there. This means that any pockets of
sorted keys are skipped over, not just if they occur at the end. Bubblesort2
code is shown in Figure \vref{Bubblesort2 code}.

@@

\section{Shakersort}
Shakersort is sometimes called back-bubblesort. It behaves like a bubblesort
which moves in both directions rather than just one way. Both ends of the array
become sorted in this manner, with the larger keys slowly moving right and the
smaller keys moving left. Sample code for shakersort is shown in Figure
\vref{Shakersort code}.

\code{Shakersort code}{on2_shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys.

The cache performance of shakersort is expected to be superior to that of
bubblesort. A small amount of temporal reuse occurs at the ends of the array,
as each key is used twice once its loaded into the cache. In addition, when the
array is not more than twice the size of the cache, then the keys in the centre
of the array are not ejected from the cache at all.

@@

%TODO check that this is put in after the titlepage
%TODO check that the captions are the same as the captions at the end of the
%other chapters
% I'm doing a slightly different result here. Level 2 doesn't interest us, so put
% in 2 prediction results instead - 1 for select/insert, 1 for bubble/shaker
\afterpage{
\label{Elementary sort results}

@@

Insertion sort performs best of these sorts in relation to cache misses. This
is due to the fact that the keys it considered most recently are used again for
the next key. These keys will be swapped out fairly regularly, however, but not
nearly as often as selection sort, where they are swapped out at nearly every
iteration.
